\documentclass[11pt, letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{environ} % Helps debug unclosed environments
\geometry{letterpaper, margin=1in}
\usepackage{times} % Using Times font for a classic brief look
\usepackage{url} % For potentially better URL handling in footnotes
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan, % Make URLs visibly distinct
    pdftitle={Fact Brief: Alleged Unconstitutional Actions by U.S. Immigration Authorities (Early 2025)},
    pdfpagemode=UseNone,
    breaklinks=true % Allows long URLs to break across lines
    }
\usepackage{parskip} % Adds space between paragraphs instead of indentation
\usepackage{footmisc} % Provides better control over footnote appearance
\usepackage{fancyhdr}
\usepackage{xcolor} % To add color, optional but recommended for visibility
\usepackage{ulem} % For strike-through/underline if needed (use sparingly)

% Configure fancyhdr
\pagestyle{fancy} % Use the 'fancy' page style
\fancyhf{} % Clear all header and footer fields
\fancyhead[C]{\textcolor{red}{\bfseries\large DRAFT - DO NOT CITE}} % Place text in Center Header
% \fancyhead[R]{\textcolor{red}{\bfseries DRAFT - DO NOT CITE}} % Alternative: Right Header
% \fancyfoot[C]{\textcolor{red}{\bfseries DRAFT - DO NOT CITE} --- \thepage} % Alternative: Footer Center with page number
\fancyfoot[C]{\thepage} % Keep page number in footer center (or choose another location)
\renewcommand{\headrulewidth}{0.4pt} % Optional: Add a rule below the header
% \renewcommand{\footrulewidth}{0.4pt} % Optional: Add a rule above the footer


\renewcommand{\footnoterule}{%
  \kern -3pt
  \hrule width \textwidth height 0.4pt
  \kern 2.6pt
} % Standard footnote separator rule

\setlength{\footnotesep}{1.5\baselineskip} % Increase spacing between footnotes for readability

\title{\textbf{Unconstitutional Actions by U.S. Immigration Authorities, Early 2025}}
\author{Joshua G. Stern}
\date{March 29, 2025} % Updated date to reflect current time

\begin{document}
\maketitle
\thispagestyle{fancy} % Apply fancy style to the first page too

\section*{Overview}

In the first quarter of 2025, under the Trump administration, U.S. federal authorities, particularly Immigration and Customs Enforcement (ICE) and the Department of Homeland Security (DHS), engaged in several high-profile immigration enforcement actions raising significant constitutional, statutory, and international law concerns.\footnote{This brief documents developments between January and March 2025, relying on the public sources cited. The velocity and contested nature of the events described – characteristic maneuvers when executive power seeks to bypass established legal constraints – necessitated rapid collation; I employed Large Language Models paired with automated searches to assemble these initial reports. I have reviewed this material and take responsibility for this draft, which I produced under conditions of deliberate speed. Recognizing the patterns herein – the invocation of emergency powers or archaic laws against specific groups, the pressure on judicial independence, the targeting of speech deemed undesirable, the attempt to create spaces beyond the reach of law (cf. authoritarian regimes) – is not merely an academic task. History provides grim precedents (e.g., Japanese-American imprisonment, partially enabled by the same Alien Enemies Act). Therefore, while this draft aims for accuracy, readers must themselves engage critically, remember that it is a draft, consult primary sources, and stay woke.} Major news organizations and legal filings report and detail instances of mass deportations bypassing due process via an archaic wartime law,\footnote{See, e.g., Al Jazeera, \emph{US deports over 200 Venezuelan immigrants to El Salvador despite court ban}, \href{https://www.aljazeera.com/news/2025/3/16/venezuelan-immigrants-deported-to-el-salvador-despite-us-court-ban}{[Source Link]} (Mar. 16, 2025); NPR, \emph{White House deports hundreds of alleged gang members to El Salvador}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=The\%20Alien\%20Enemies\%20Act\%20allows,members\%20of\%20a\%20criminal\%20gang}{[Source Link]} (Mar. 16, 2025). The Alien Enemies Act (AEA) of 1798 allows summary removal of citizens of hostile nations during declared war or invasion, bypassing normal Immigration and Nationality Act (INA) procedures.} blatant defiance of federal court orders halting those deportations,\footnote{See, e.g., The Guardian, \emph{US deports 250 alleged gang members to El Salvador despite court ruling to halt flights}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=The\%20US\%20district\%20judge\%20James,to\%20return\%20to\%20the\%20US}{[Source Link]} (Mar. 16, 2025); Associated Press (AP), \emph{Trump administration deports hundreds of immigrants, despite court order}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=Boasberg\%20barred\%20those\%20Venezuelans\%E2\%80\%99\%20deportations,legal\%20authority\%20in\%20invoking\%20it}{[Source Link]} (Mar. 16, 2025); CBS News, \emph{Trump administration invokes state secrets privilege in deportation case}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=federal\%20courts}{[Source Link]} (Mar. 24, 2025) (reporting administration ignored court's TRO and later invoked state secrets privilege to avoid judicial inquiry).} targeted detentions seemingly aimed at chilling political speech, particularly against pro-Palestinian activists, raising First Amendment concerns,\footnote{See, e.g., The Guardian, \emph{‘A warning for students of color’: Ice agents are targeting certain protesters, say experts}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=Now\%2C\%20at\%20least\%20five\%20students,over\%20their\%20support\%20of\%20Palestine}{[Source Link]} (Mar. 26, 2025); Al Jazeera, \emph{Who are the students Trump wants to deport?}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport}{[Source Link]} (Mar. 27, 2025); CBS News, \emph{ICE detains University of Alabama doctoral student}, \href{https://www.cbsnews.com/news/ice-detains-university-of-alabama-doctoral-student/?intcid=CNI-00-10aaa3a\#:~:text=Mahmoud\%20Khalil\%20\%27s\%20case\%20has,terrorist\%20sympathizers}{[Source Link]} (Mar. 27, 2025) (reporting Trump labeled detained activist a "terrorist sympathizer" linking detention to protest activity).} and plans for extraordinary transfers of detainees to offshore locations like Guantánamo Bay, potentially violating habeas corpus rights and seeking to place individuals beyond the reach of U.S. courts.\footnote{See CBS News, \emph{Trump administration sends first group of migrant detainees to Guantanamo Bay}, \href{https://www.cbsnews.com/news/trump-guantanamo-bay-migrant-detainees/}{[Source Link]} (Feb. 5, 2025); AP, \emph{After a stint in Guantanamo Bay, a Venezuelan deported from the US adjusts to his homeland}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=Bastidas\%20and\%20roughly\%20350\%20other,government\%20of\%20President\%20Nicol\%C3\%A1s\%20Maduro}{[Source Link]} (Mar. 25, 2025) (reporting ~180 Venezuelans held at Gitmo for up to 16 days); Cf. \textit{Boumediene v. Bush}, 553 U.S. 723 (2008) (affirming habeas rights for Gitmo detainees).} Further concerns involve Fourth Amendment violations through warrantless ICE raids and secretive detention tactics.\footnote{ACLU of Illinois Press Release, \emph{22 People Arrested in ICE Raids...}, \href{https://www.aclu-il.org/en/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=The\%20National\%20Immigrant\%20Justice\%20Center,without\%20warrants\%20or\%20probable\%20cause}{[Source Link]} (Mar. 17, 2025) (reporting warrantless arrests violating prior settlement).}

\section*{Mass Deportation of Venezuelans to El Salvador (March 2025)}

\subsection*{The Operation}
In mid-March 2025, a U.S. operation transferred approximately \textbf{238 Venezuelan men} to El Salvador for imprisonment.\footnote{CBS News, \emph{Here are the names of the Venezuelans deported by the U.S. to El Salvador}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=On\%20March\%2015\%2C\%20the\%20U,terrorist\%20group\%20and\%20wartime\%20enemy}{[Source Link]} (reporting 238 Venezuelan men on three charter flights March 15); The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=El\%20Salvador\%E2\%80\%99s\%20president\%2C\%20Nayib\%20Bukele\%2C,person\%20capacity\%20\%E2\%80\%9Cterrorism\%20confinement\%20centre\%E2\%80\%9D}{[Source Link]} (reporting Bukele announced 238 Venezuelans arrived alongside Salvadoran gang members, totaling over 250); Al Jazeera, \emph{Venezuela resumes accepting people deported from US}, \href{https://www.aljazeera.com/news/2025/3/24/venezuela-resumes-accepting-people-deported-from-us\#:~:text=Washington\%20then\%20deported\%20238\%20Venezuelans,security\%20prison\%20in\%20El\%20Salvador}{[Source Link]} (Mar. 24, 2025).} Prior to this transfer, reports indicate around 180 Venezuelans had been detained for up to 16 days in makeshift holding tents at the U.S. Naval Base at Guantánamo Bay, Cuba.\footnote{AP, \emph{After a stint in Guantanamo Bay...}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=Bastidas\%20and\%20roughly\%20350\%20other,government\%20of\%20President\%20Nicol\%C3\%A1s\%20Maduro}{[Source Link]} (Mar. 25, 2025); Reuters, \emph{Trump deporting people...}, \href{https://www.reuters.com/world/us/trump-set-broaden-arrests-deportation-routes-expand-immigration-crackdown-2025-02-21/\#:~:text=Item\%201\%20of\%203\%20Newly,Bay\%20Public\%20Affairs\%2FHandout\%20via\%20REUTERS}{[Source Link]} (Feb 21, 2025) (showing images of holding tents).} On March 15, President Trump issued a proclamation invoking the rarely used \textbf{Alien Enemies Act (AEA) of 1798} (50 U.S.C. §§ 21–24),\footnote{NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=It\%20was\%20not\%20immediately\%20clear,territory}{[Source Link]} (Mar. 16, 2025); The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=The\%20US\%20deported\%20more\%20than,to\%20be\%20used\%20in\%20wartime}{[Source Link]} (Mar. 16, 2025); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=In\%20a\%20proclamation\%C2\%A0invoking\%20the\%20Alien,and\%20removed\%20as\%20Alien\%20Enemies}{[Source Link]} (Mar. 28, 2025).} last significantly used during World War II,\footnote{CBS News, \emph{Trump administration invokes state secrets...}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=panel\%20didn\%27t\%20ask\%20any\%20questions}{[Source Link]} (Mar. 24, 2025).} designating members of the Venezuelan gang Tren de Aragua as "enemy aliens" involved in a "predatory incursion," thereby authorizing their summary removal.\footnote{NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=The\%20Alien\%20Enemies\%20Act\%20allows,members\%20of\%20a\%20criminal\%20gang}{[Source Link]}. Tren de Aragua is described as a gang formed in Venezuelan prisons. See CBS News, \emph{Here are the names...}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=The\%20Trump\%20administration\%20has\%20said,that\%20while\%20some\%20of\%20those}{[Source Link]}. Relatives deny some deportees are members, and little evidence was presented publicly. AP, \emph{After a stint in Guantanamo Bay...}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=It\%20is\%20part\%20of\%20the,evidence\%20to\%20back\%20that\%20up}{[Source Link]}. CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=In\%20a\%20proclamation\%C2\%A0invoking\%20the\%20Alien,and\%20removed\%20as\%20Alien\%20Enemies}{[Source Link]} (Mar. 28, 2025).} This allowed for \textbf{expedited removal without standard immigration court hearings or asylum claims}\footnote{NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=immigration\%20proceedings\%20in\%20court\%2C\%20or,members\%20of\%20a\%20criminal\%20gang}{[Source Link]} ("would not go through the normal immigration proceedings in court, or be able to claim asylum"); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=match\%20at\%20L234\%20Alien\%20Enemies,Act\%2C\%20which\%20precludes\%20judicial\%20review}{[Source Link]} (noting the AEA "precludes judicial review").} for at least 137 of the men; another 101 were reportedly deported under regular INA procedures but amid the same operation.\footnote{CBS News, \emph{Here are the names...}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=U,procedures\%2C\%20the\%20officials\%20have\%20said}{[Source Link]} (reporting U.S. official clarification on numbers); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=However\%2C\%20on\%20the\%20same\%20day,the\%20nation\%27s\%20maximum\%20security\%20prison}{[Source Link]} (confirming 137 under AEA).} The men were flown on charter flights on March 15, 2025,\footnote{CBS News, \emph{Here are the names...}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=On\%20March\%2015\%2C\%20the\%20U,terrorist\%20group\%20and\%20wartime\%20enemy}{[Source Link]}. Note: While flights departed or were in process on March 15, arrival and processing in El Salvador extended into March 16.} handed over to Salvadoran guards, and imprisoned in the CECOT "mega-prison,"\footnote{CBS News, \emph{Here are the names...}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=group\%20and\%20wartime\%20enemy}{[Source Link]} (describing handover, head shaving, march into CECOT).} known for housing gang members, under an agreement where the U.S. pays El Salvador for their detention.\footnote{The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=El\%20Salvador\%E2\%80\%99s\%20president\%2C\%20Nayib\%20Bukele\%2C,person\%20capacity\%20\%E2\%80\%9Cterrorism\%20confinement\%20centre\%E2\%80\%9D}{[Source Link]}; Reuters, \emph{El Salvador Supreme Court to hear defense of jailed Venezuelans...}, \href{https://www.reuters.com/world/americas/lawyers-defend-30-venezuelans-deported-us-el-salvadors-supreme-court-2025-03-24/\#:~:text=law\%20targeting\%20,Venezuelan\%20citizens\%20to\%20turn\%20around}{[Source Link]} (Mar. 24, 2025).}

\subsection*{Bypassing Due Process and Statutory Authority}
The use of the Alien Enemies Act was explicitly intended to circumvent normal INA immigration proceedings and associated due process protections.\footnote{See NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=immigration\%20proceedings\%20in\%20court\%2C\%20or,members\%20of\%20a\%20criminal\%20gang}{[Source Link]}.} Reports confirm these individuals, many allegedly asylum seekers or without criminal convictions,\footnote{AP, \emph{Trump administration deports hundreds...}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=match\%20at\%20L1527\%20any\%20evidence,arrested\%20in\%20the\%20United\%20States}{[Source Link]} (Mar. 16, 2025) (noting advocates argue many were asylum seekers or not convicted); AP, \emph{Trump administration deports hundreds...}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=contending\%20Tren\%20de\%20Aragua\%20was,They\%20began\%20to\%20file}{[Source Link]} (Mar. 16, 2025).} were denied the opportunity to contest the gang affiliation claims or seek protection in U.S. courts, violating core tenets of procedural due process guaranteed by the Fifth Amendment.\footnote{PBS NewsHour, \emph{ACLU attorney on suing Trump administration...}, \href{https://www.pbs.org/newshour/show/aclu-attorney-on-suing-trump-administration-for-using-wartime-authority-for-deportations\#:~:text=President\%20Trump\%E2\%80\%99s\%20use\%20of\%20the,the\%20ACLU\%27s\%20Immigrants\%20Rights\%20Project}{[Source Link]} (Mar. 18, 2025) (stating deportations occurred "without due process"); ACLU Press Release, \emph{ACLU and Democracy Forward Comment...}, \href{https://www.aclu.org/press-releases/aclu-foia-litigation-continues-to-disclose-ice-proposals-to-expand-immigration-detention-nationwide\#:~:text=ACLU\%20and\%20Democracy\%20Forward\%20Comment,in\%20Alien\%20Enemies\%20Act\%20Challenge}{[Source Link]} (Mar. 15, 2025) (quoting Lee Gelernt on lack of process). Judge Patricia Millett noted the lack of process was starker than that afforded Nazi detainees in WWII. CBS News, \emph{Appeals court upholds order...}, \href{https://www.cbsnews.com/news/appeals-court-stopping-deportations-alien-enemies-act/?intcid=CNI-00-10aaa3a\#:~:text=Judge\%20Millett\%2C\%20who\%20was\%20appointed,tortured\%20and\%20their\%20lives\%20endangered}{[Source Link]} (Mar. 26, 2025).} This action also likely violated the Refugee Act of 1980 (implementing U.S. obligations under the 1967 Refugee Protocol and Convention Against Torture) by failing to screen for asylum or fear of torture (non-refoulement principle).\footnote{U.S. law, e.g., 8 U.S.C. § 1158, § 1231(b)(3), requires protection screening. Mass expulsion without screening likely violates these statutes and the international principle of non-refoulement. Subsequent reports indicated Venezuela was imprisoning some returnees. AP, \emph{After a stint in Guantanamo Bay...}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=It\%20is\%20part\%20of\%20the,evidence\%20to\%20back\%20that\%20up}{[Source Link]}.} Furthermore, the legal basis for invoking the AEA was highly questionable, as the Act requires a "declared war" or "invasion" by a foreign nation, conditions arguably not met by the activities of a criminal gang, rendering the action potentially \textit{ultra vires} (beyond statutory authority).\footnote{Judge Boasberg questioned this premise. CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=use\%20in\%20this\%20case}{[Source Link]} (Mar. 28, 2025). Invoking the AEA against a non-state actor without Congressional declaration of war stretches the statute's scope.} Critics, including the ACLU, argue this represents a major violation of statutory and constitutional due process rights.\footnote{ACLU Press Release, \emph{ACLU and Democracy Forward Comment...}, \href{https://www.aclu.org/press-releases/aclu-foia-litigation-continues-to-disclose-ice-proposals-to-expand-immigration-detention-nationwide\#:~:text=request\%20to\%20lift\%20a\%20TRO,He}{[Source Link]} (Mar. 15, 2025) (describing the action as "unprecedented and unlawful").}

\subsection*{Legal Challenge and Defiance of Court Order}
The ACLU filed an emergency lawsuit on March 15, 2025.\footnote{AP, \emph{Trump administration deports hundreds...}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=The\%20litigation\%20that\%20led\%20to,remove\%20them\%20from\%20the\%20country}{[Source Link]} (Mar. 16, 2025); WDSU News (AP wire), \emph{Trump seeks Supreme Court nod on Venezuelan deportations}, \href{https://www.wdsu.com/article/trump-supreme-court-venezuelan-deportations/64321685\#:~:text=organizations\%20and\%20risk\%20debilitating\%20effects,vowed\%20to\%20determine\%20whether\%20the}{[Source Link]} (Mar. 28, 2025).} U.S. District Chief Judge \textbf{James E. Boasberg} issued a temporary restraining order (TRO) that evening, blocking the deportations under the Alien Enemies Act for the plaintiffs and later broadening it to cover all targeted individuals.\footnote{AP, \emph{Trump administration deports hundreds...}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=Boasberg\%20barred\%20those\%20Venezuelans\%E2\%80\%99\%20deportations,legal\%20authority\%20in\%20invoking\%20it}{[Source Link]} (Mar. 16, 2025); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=quickly\%20kicked\%20off\%20a\%20legal,battle}{[Source Link]}.}. Judge Boasberg explicitly ordered, including via oral instruction from the bench, any planes already airborne carrying individuals under the AEA to be turned around.\footnote{The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=The\%20US\%20district\%20judge\%20James,to\%20return\%20to\%20the\%20US}{[Source Link]} (Mar. 16, 2025); AP, \emph{Trump administration deports hundreds...}, \href{https://apnews.com/article/trump-venezuela-el-salvador-immigration-dd4f61999f85c4dd8bcaba7d4fc7c9af\#:~:text=He\%20said\%20he\%20had\%20to,their\%20pleas\%20heard\%20in\%20court}{[Source Link]} (Mar. 16, 2025); CBS News, \emph{Trump administration invokes state secrets...}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=U,administration\%20and\%20the\%20federal\%20courts}{[Source Link]}.}.

However, the deportation flights proceeded, landing in El Salvador on March 16.\footnote{NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=White\%20House\%20deports\%20hundreds\%20of,any\%20planes\%20in\%20the\%20air}{[Source Link]} (Mar. 16, 2025); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=However\%2C\%20on\%20the\%20same\%20day,the\%20nation\%27s\%20maximum\%20security\%20prison}{[Source Link]} (confirming three flights landed).} The administration argued the TRO came "too late," after the planes had left U.S. airspace, and disputed the binding nature of the oral order.\footnote{NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=It\%20was\%20not\%20immediately\%20clear,territory}{[Source Link]} (Mar. 16, 2025) (citing officials claiming order came too late); NPR, \emph{White House deports hundreds...}, \href{https://www.npr.org/2025/03/16/g-s1-54154/alien-enemies-el-salvador-trump\#:~:text=members\%20subject\%20to\%20removal\%20under,Deputy\%20Attorney\%20General\%20Todd\%20Blanche}{[Source Link]} (citing govt lawyers' statement that some were removed before the 7:26 PM filing); CBS News, \emph{Trump administration invokes state secrets...}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=During\%20a\%20hearing\%20Friday\%2C\%20Boasberg,S}{[Source Link]} (reporting DOJ argued oral order "did not count").} A White House statement asserted that a single judge could not compel the return of aircraft carrying "foreign alien terrorists."\footnote{The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=\%E2\%80\%9CA\%20single\%20judge\%20in\%20a,Leavitt\%20said\%20in\%20a\%20statement}{[Source Link]} (Mar. 16, 2025) (quoting White House statement).} Reports confirm the flights were not recalled, and the men were delivered to El Salvador \textbf{despite the judicial injunction},\footnote{Reuters, \emph{El Salvador Supreme Court to hear defense...}, \href{https://www.reuters.com/world/americas/lawyers-defend-30-venezuelans-deported-us-el-salvadors-supreme-court-2025-03-24/\#:~:text=Some\%20137\%20of\%20the\%20group,Venezuelan\%20citizens\%20to\%20turn\%20around}{[Source Link]} (Mar. 24, 2025); CBS News, \emph{Here are the names...}, \href{https://www.cbsnews.com/news/venezuelans-deported-el-salvador-names/\#:~:text=The\%20high,and\%20turn\%20the\%20flights\%20around}{[Source Link]}; The Guardian, \emph{US deports 250 alleged gang members...}, \href{https://www.theguardian.com/us-news/2025/mar/16/deportation-alleged-gang-members-el-salvador\#:~:text=air\%20to\%20return\%20to\%20the,US}{[Source Link]} (Mar. 16, 2025) (reporting Bukele's tweet confirming arrival despite order); AP, \emph{After a stint in Guantanamo Bay...}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=Over\%20the\%20weekend\%2C\%20the\%20U,membership\%20claim}{[Source Link]} (noting deportations occurred "even as a federal judge issued an order temporarily barring" them).} an action undermining the rule of law and judicial independence (Article III). Judge Boasberg subsequently vowed to investigate whether the government unlawfully defied his order, though the administration invoked the extraordinary \textbf{state secrets privilege} to avoid disclosing details about the flights, hindering judicial review.\footnote{WDSU News (AP wire), \emph{Trump seeks Supreme Court nod...}, \href{https://www.wdsu.com/article/trump-supreme-court-venezuelan-deportations/64321685\#:~:text=proclamation\%20was\%20made\%20public,his\%20allies\%20have\%20called\%20for}{[Source Link]} (Mar. 28, 2025) (reporting Boasberg's vow and state secrets claim); CBS News, \emph{Trump administration invokes state secrets...}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=The\%20Trump\%20administration\%20invoked\%20a,El\%20Salvador\%20under\%20war\%20authorities}{[Source Link]} (Mar. 24, 2025). The privilege claim, rare in immigration, was seen as an attempt to shield executive actions from accountability.} President Trump and allies publicly attacked Judge Boasberg, calling for his impeachment and disbarment, drawing a rebuke from Chief Justice John Roberts defending judicial independence.\footnote{CBS News, \emph{Trump administration invokes state secrets...}, \href{https://www.cbsnews.com/colorado/news/trump-state-secrets-privilege-deportations-alien-enemies-act/?intcid=CNR-01-0623\#:~:text=,the\%20judge\%20wrote}{[Source Link]} (Mar. 24, 2025).}

\subsection*{Ongoing Legal Battle}
The administration appealed Judge Boasberg's TRO.\footnote{Reuters, \emph{US appeals court upholds block on Trump deportation...}, \href{https://www.reuters.com/world/us/us-homeland-secretary-visit-el-salvador-prison-holding-deported-venezuelans-2025-03-26/\#:~:text=Venezuelan\%20immigrants\%20under\%20a\%20little,18th\%20century\%20law}{[Source Link]} (Mar. 26, 2025); WDSU News (AP wire), \emph{Trump seeks Supreme Court nod...}, \href{https://www.wdsu.com/article/trump-supreme-court-venezuelan-deportations/64321685\#:~:text=Enemies\%20Act\%20for\%20the\%20first,Tuesday\%20deadline\%20for\%20a\%20response}{[Source Link]} (Mar. 28, 2025).} On March 26, 2025, the D.C. Circuit Court of Appeals, in a 2-1 decision, \textbf{upheld the injunction} blocking further use of the Alien Enemies Act for these deportations, emphasizing the lack of due process.\footnote{Reuters, \emph{US appeals court upholds block...}, \href{https://www.reuters.com/world/us/us-homeland-secretary-visit-el-salvador-prison-holding-deported-venezuelans-2025-03-26/\#:~:text=WASHINGTON\%2C\%20March\%2026\%20\%28Reuters\%29\%20\%20,used\%2018th\%20century\%20law}{[Source Link]} (Mar. 26, 2025); Reuters, \emph{US appeals court upholds block...}, \href{https://www.reuters.com/world/us/us-homeland-secretary-visit-el-salvador-prison-holding-deported-venezuelans-2025-03-26/\#:~:text=A\%20three,dissenting}{[Source Link]} (Mar. 26, 2025) (noting 2-1 vote); CBS News, \emph{Appeals court upholds order...}, \href{https://www.cbsnews.com/news/appeals-court-stopping-deportations-alien-enemies-act/?intcid=CNI-00-10aaa3a\#:~:text=Judges\%20Karen\%20LeCraft\%20Henderson\%20and,favor\%20of\%20the\%20Justice\%20Department}{[Source Link]} (Mar. 26, 2025) (listing Judges Henderson (Reagan appointee) and Millett (Obama appointee) in majority, Walker (Trump appointee) dissenting).} The administration subsequently petitioned the U.S. Supreme Court on March 28, 2025, seeking to lift the ban.\footnote{WDSU News (AP wire), \emph{Trump seeks Supreme Court nod...}, \href{https://www.wdsu.com/article/trump-supreme-court-venezuelan-deportations/64321685\#:~:text=By\%20MARK\%20SHERMAN\%20Associated\%20Press}{[Source Link]} (Mar. 28, 2025); WDSU News (AP wire), \emph{Trump seeks Supreme Court nod...}, \href{https://www.wdsu.com/article/trump-supreme-court-venezuelan-deportations/64321685\#:~:text=the\%20American\%20Civil\%20Liberties\%20Union,his\%20allies\%20have\%20called\%20for}{[Source Link]} (Mar. 28, 2025) (noting SCOTUS required ACLU response); CBS News, \emph{Trump's use of the Alien Enemies Act...}, \href{https://www.cbsnews.com/news/trump-alien-enemies-act-of-1798-history/?intcid=CNR-01-0623\#:~:text=federal\%20appeals\%20court\%20\%20on,lift\%20the\%20lower\%20court\%20orders}{[Source Link]}. As of this writing, the injunction remains in place.}

\section*{Targeted Detentions and Raids (Jan–Mar 2025)}

\subsection*{Crackdown on Pro-Palestinian Student Activists}
A distinct pattern emerged of ICE targeting international students and academics involved in campus protests against the Gaza war, raising serious First Amendment concerns about retaliation for protected speech and viewpoint discrimination.\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=Now\%2C\%20at\%20least\%20five\%20students,over\%20their\%20support\%20of\%20Palestine}{[Source Link]} (Mar. 26, 2025) (reporting at least five students targeted); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport}{[Source Link]} (Mar. 27, 2025) (detailing six cases). Targeting non-citizens for deportation based on their political expression violates the First Amendment. See, e.g., \textit{Bridges v. Wixon}, 326 U.S. 135 (1945).} At least six individuals faced visa revocation, detention, and/or deportation threats, actions widely condemned by civil liberties groups as unprecedented and unconstitutional punishment for political speech:\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=The\%20targeting\%20of\%20immigrant\%20students,\%E2\%80\%9D}{[Source Link]} (Mar. 26, 2025) (quoting experts calling it "alarming escalation" and "unprecedented"); RFK Human Rights, \emph{Justice Roundup- March 2025}, \href{https://www.linkedin.com/pulse/justice-roundup-march-2025-rfkhumanrights-glnxe\#:~:text=The\%20Central\%20Louisiana\%20ICE\%20Processing,light\%20on\%20egregious\%20human\%20rights}{[Source Link]} (Mar. 2025) (calling detentions an attempt to silence dissent).}
    \begin{itemize}
        \item \textbf{Mahmoud Khalil (Columbia):} A \textbf{lawful permanent resident (LPR)} arrested by ICE without warrant (March 8),\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=On\%20March\%208\%2C\%20ICE\%20agents,kind\%20under\%20the\%20Trump\%20administration}{[Source Link]} (Mar. 27, 2025); ACLU Press Release, \emph{Court Rules Mahmoud Khalil’s Lawsuit...}, \href{https://www.aclu.org/press-releases/court-rules-mahmoud-khalils-lawsuit-challenging-his-unlawful-detention-by-ice-should-move-forward-in-new-jersey\#:~:text=Samah\%20Sisay\%2C\%20staff\%20attorney\%20at,\%E2\%80\%9D}{[Source Link]} (Mar. 19, 2025) (noting wife reported no warrant shown).} transferred overnight to Louisiana.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=McLaughlin\%20alleged\%20Khalil\%20\%E2\%80\%9Cled\%20activities,processing\%20facility\%20in\%20Jena\%2C\%20Louisiana}{[Source Link]} (Mar. 27, 2025).} Accused of Hamas ties without public evidence; President Trump publicly labeled him a "terrorist sympathizer" whose deportation he supported, strongly suggesting retaliatory motive for protest activity.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=McLaughlin\%20alleged\%20Khalil\%20\%E2\%80\%9Cled\%20activities,processing\%20facility\%20in\%20Jena\%2C\%20Louisiana}{[Source Link]} (Mar. 27, 2025); Reuters, \emph{Judge rules pro-Palestinian student...}, \href{https://www.reuters.com/legal/judge-rules-pro-palestinian-student-permanent-us-resident-cannot-be-detained-2025-03-25/\#:~:text=Columbia\%20protester\%20Mahmoud\%20Khalil\%2C\%20who,supporting\%20Hamas\%2C\%20which\%20Khalil\%20denies}{[Source Link]} (Mar. 25, 2025) (noting Khalil denies claim); CBS News, \emph{ICE detains University of Alabama doctoral student...}, \href{https://www.cbsnews.com/news/ice-detains-university-of-alabama-doctoral-student/?intcid=CNI-00-10aaa3a\#:~:text=Mahmoud\%20Khalil\%20\%27s\%20case\%20has,terrorist\%20sympathizers}{[Source Link]} (Mar. 27, 2025).} Court blocked immediate deportation and returned case venue to NY/NJ.\footnote{ACLU Press Release, \emph{Court Rules Mahmoud Khalil’s Lawsuit...}, \href{https://www.aclu.org/press-releases/court-rules-mahmoud-khalils-lawsuit-challenging-his-unlawful-detention-by-ice-should-move-forward-in-new-jersey\#:~:text=NEW\%20YORK\%20\%E2\%80\%93\%20The\%20Southern,absence\%20of\%20a\%20court\%20order}{[Source Link]} (Mar. 19, 2025).}
        \item \textbf{Rumeysa Ozturk (Tufts):} Turkish PhD student (likely on F-1 visa) seized by plainclothes agents in unmarked cars (March 25),\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Security\%20camera\%20footage\%20from\%20Tuesday,to\%20break\%20her\%20Ramadan\%20fast}{[Source Link]} (Mar. 27, 2025) (reporting seizure near her home); Tufts Daily article cited in user prompt: \href{https://www.tuftsdaily.com/article/2024/03/4ftk27sm6jkj}{[Source Link]}.} described as an "abduction."\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=match\%20at\%20L186\%20The\%20video,They\%E2\%80\%99re\%20in\%20unmarked}{[Source Link]} (Mar. 27, 2025) (quoting witness description).} Transferred overnight to Louisiana, potentially in defiance of a court order requiring notice before out-of-state transfer.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Ozturk\%E2\%80\%99s\%20lawyer\%2C\%20Mahsa\%20Khanbabai\%2C\%20filed,Massachusetts\%20without\%2048\%20hours\%20notice}{[Source Link]} (Mar. 27, 2025) (reporting lawyer obtained order, but ICE transferred her that night); court order: \href{https://lnkd.in/eWNae6hZ}{[Source Link]}.} DHS accused her of supporting Hamas, again without public evidence.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=match\%20at\%20L168\%20an\%20X,specify\%20what\%20these\%20activities\%20were}{[Source Link]} (Mar. 27, 2025) (quoting DHS statement on social media); CBS News, \emph{ICE detains University of Alabama doctoral student...}, \href{https://www.cbsnews.com/news/ice-detains-university-of-alabama-doctoral-student/?intcid=CNI-00-10aaa3a\#:~:text=A\%20Tufts\%20University\%20graduate\%20student,details\%20about\%20her\%20alleged\%20activities}{[Source Link]} (Mar. 27, 2025).}
        \item \textbf{Badar Khan Suri (Georgetown):} Indian postdoctoral fellow arrested (March 17)\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Badar\%20Khan\%20Suri}{[Source Link]} (Mar. 27, 2025).} invoking a rare "adverse foreign policy consequences" clause (INA § 212(a)(3)(C)).\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=In\%20the\%20case\%20of\%20Khalil,promoting\%20antisemitism\%20on\%20social\%20media\%E2\%80\%9D}{[Source Link]} (Mar. 26, 2025).} Accused of spreading Hamas propaganda.\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=activism\%2C\%20the\%20administration\%20cited\%20a,promoting\%20antisemitism\%20on\%20social\%20media\%E2\%80\%9D}{[Source Link]} (Mar. 26, 2025); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=McLaughlin\%20attributed\%20Suri\%E2\%80\%99s\%20arrest\%20to,\%E2\%80\%9D}{[Source Link]} (Mar. 27, 2025).} Judge blocked deportation; held in Texas.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=However\%2C\%20the\%20Trump\%20administration\%20has,Semitism}{[Source Link]} (Mar. 27, 2025) (reporting judge blocked deportation Mar. 20, Suri remains in custody).}
        \item \textbf{Yunseo Chung (Columbia):} South Korean green card holder (LPR since age 7) targeted by ICE after NYPD protest arrest (March 5).\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=Yunseo\%20Chung\%2C\%20a\%2021,until\%20the\%20court\%20rules\%20otherwise}{[Source Link]} (Mar. 26, 2025); Reuters, \emph{Judge rules pro-Palestinian student...}, \href{https://www.reuters.com/legal/judge-rules-pro-palestinian-student-permanent-us-resident-cannot-be-detained-2025-03-25/\#:~:text=The\%20U,Hamas}{[Source Link]} (Mar. 25, 2025).} Went into hiding; judge blocked ICE from detaining her (March 25), affirming LPRs cannot be held for protected speech.\footnote{The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=Yunseo\%20Chung\%2C\%20a\%2021,until\%20the\%20court\%20rules\%20otherwise}{[Source Link]} (Mar. 26, 2025); Reuters, \emph{Judge rules pro-Palestinian student...}, \href{https://www.reuters.com/legal/judge-rules-pro-palestinian-student-permanent-us-resident-cannot-be-detained-2025-03-25/\#:~:text=U,being\%20detained\%2C\%20court\%20records\%20showed}{[Source Link]} (Mar. 25, 2025).}
        \item \textbf{Momodou Taal (Cornell):} Gambian/UK PhD student had visa revoked (March 14).\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Advertisement}{[Source Link]} (Mar. 27, 2025); Reuters, \emph{Judge rules pro-Palestinian student...}, \href{https://www.reuters.com/legal/judge-rules-pro-palestinian-student-permanent-us-resident-cannot-be-detained-2025-03-25/\#:~:text=Badar\%20Khan\%20Suri\%20\%2C\%20an,A\%20federal\%20judge\%20\%20102}{[Source Link]} (Mar. 25, 2025).} Faced ICE pressure to self-surrender; joined lawsuit challenging targeting of protesters.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=On\%20the\%20morning\%20of\%20March,his\%20residence\%2C\%20including\%20on\%20campus}{[Source Link]} (Mar. 27, 2025); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Taal\%20told\%20Al\%20Jazeera\%20that,one\%20focused\%20on\%20university\%20campuses}{[Source Link]} (Mar. 27, 2025).}
        \item \textbf{Ranjani Srinivasan (Columbia):} Indian PhD candidate had visa revoked (March 5).\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Ranjani\%20Srinivasan\%2C\%2037\%2C\%20had\%20her,a\%20valid\%20visa\%20until\%202029}{[Source Link]} (Mar. 27, 2025).} Fled U.S. for Canada (March 11) after ICE visit and fearing detention.\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=On\%20March\%207\%2C\%20individuals\%20claiming,the\%20US\%2C\%20before\%20eventually\%20leaving}{[Source Link]} (Mar. 27, 2025); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=After\%20this\%2C\%20Srinivasan\%20left\%20her,\%E2\%80\%9D}{[Source Link]} (Mar. 27, 2025).} DHS Secretary Kristi Noem publicly posted video of her departure, labeling her a "terrorist sympathizer."\footnote{Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=On\%20March\%2014\%2C\%20DHS\%20Secretary,sympathizers\%E2\%80\%9D\%20leaving\%20on\%20their\%20own}{[Source Link]} (Mar. 27, 2025).}
        \item \textbf{Alireza Doroudi (Univ. Alabama):} Iranian doctoral student detained in pre-dawn raid with no public explanation, exemplifying climate of fear.\footnote{CBS News, \emph{ICE detains University of Alabama doctoral student...}, \href{https://www.cbsnews.com/news/ice-detains-university-of-alabama-doctoral-student/?intcid=CNI-00-10aaa3a\#:~:text=Alireza\%20Doroudi\%2C\%20a\%20doctoral\%20student,his\%20home\%2C\%20the\%20paper\%20reported}{[Source Link]} (Mar. 27, 2025).}
    \end{itemize}
In these cases, authorities often cited national security or support for terrorism (presumably under INA § 212(a)(3)(B) or § 237(a)(4)(B)) but provided little public evidence, relying instead on broad immigration statutes, suggesting pretextual use of law to punish speech.\footnote{See, e.g., NPR, \emph{ICE arrests, seeks to deport Palestinian protester...}, \href{https://www.npr.org/2025/03/10/nx-s1-5323166/arrest-green-card-palestinian-protest\#:~:text=The\%20arrest\%20follows\%20through\%20on,endorses\%20or\%20espouses\%20terrorist\%20activity}{[Source Link]} (Mar. 10, 2025) (noting use of INA § 212(a)(3)(B)); The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=In\%20the\%20case\%20of\%20Khalil,promoting\%20antisemitism\%20on\%20social\%20media\%E2\%80\%9D}{[Source Link]} (Mar. 26, 2025) (noting use of INA § 212(a)(3)(C) - foreign policy consequences clause). Mere advocacy generally does not meet the statutory definition of material support for terrorism.} The use of remote detention facilities (like the GEO Group center in Jena, Louisiana)\footnote{RFK Human Rights, \emph{Justice Roundup- March 2025}, \href{https://www.linkedin.com/pulse/justice-roundup-march-2025-rfkhumanrights-glnxe\#:~:text=The\%20Central\%20Louisiana\%20ICE\%20Processing,light\%20on\%20egregious\%20human\%20rights}{[Source Link]} (Mar. 2025) (highlighting the facility's role and history).} and secretive interstate transfers\footnote{See ACLU Press Release, \emph{Court Rules Mahmoud Khalil’s Lawsuit...}, \href{https://www.aclu.org/press-releases/court-rules-mahmoud-khalils-lawsuit-challenging-his-unlawful-detention-by-ice-should-move-forward-in-new-jersey\#:~:text=Mr,rewarding\%20the\%20Trump\%20administration\%E2\%80\%99s\%20unlawful}{[Source Link]} (Mar. 19, 2025) (describing Khalil's transfer as attempt to manipulate jurisdiction); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=Ozturk\%E2\%80\%99s\%20lawyer\%2C\%20Mahsa\%20Khanbabai\%2C\%20filed,Massachusetts\%20without\%2048\%20hours\%20notice}{[Source Link]} (Mar. 27, 2025) (describing Ozturk's transfer).} has been criticized as tactics to isolate detainees, impede legal access, and violate due process rights, including the right to counsel.

\subsection*{Warrantless Raids and Secretive Tactics}
Beyond campuses, reports indicate broader potentially unlawful tactics violating the Fourth Amendment. In late January 2025, ICE raids in the Chicago area reportedly swept up 22 people, including a U.S. citizen, \textbf{without judicial warrants or probable cause}, violating a prior federal court settlement (\textit{Castañon Nava v. ICE}) according to the ACLU of Illinois and the National Immigrant Justice Center (NIJC).\footnote{ACLU of Illinois Press Release, \emph{22 People Arrested in ICE Raids...}, \href{https://www.aclu-il.org/en/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=CHICAGO\%20,a\%202022\%20federal\%20court\%20settlement}{[Source Link]} (Mar. 17, 2025); NIJC, \emph{22 People Arrested in ICE Raids Announce Federal Court Action...}, \href{https://immigrantjustice.org/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=The\%20National\%20Immigrant\%20Justice\%20Center,without\%20warrants\%20or\%20probable\%20cause}{[Source Link]} (Mar. 17, 2025).} Examples include seizing 12 people in a Missouri restaurant raid with no warrants and arresting a U.S. citizen who was then held for 10 hours without the arrest being documented, a form of "secret detention."\footnote{NIJC, \emph{22 People Arrested...}, \href{https://immigrantjustice.org/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=match\%20at\%20L105\%20who\%20have,without\%20warrants\%20or\%20probable\%20cause}{[Source Link]} (restaurant raid); NIJC, \emph{22 People Arrested...}, \href{https://immigrantjustice.org/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=The\%20National\%20Immigrant\%20Justice\%20Center,without\%20warrants\%20or\%20probable\%20cause}{[Source Link]} (U.S. citizen detention).} These "collateral arrests" were part of mass sweeps criticized by legal advocates as "lawless and reckless" and "blatantly unconstitutional."\footnote{ACLU of Illinois Press Release, \emph{22 People Arrested in ICE Raids...}, \href{https://www.aclu-il.org/en/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=\%E2\%80\%9CThe\%20Trump\%20Administration\%20once\%20again,\%E2\%80\%9D}{[Source Link]} (Mar. 17, 2025); NIJC, \emph{22 People Arrested...}, \href{https://immigrantjustice.org/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=Deportations}{[Source Link]}.} The administration also rescinded policies protecting "sensitive locations" like schools and churches from enforcement actions, broadening ICE's operational latitude.\footnote{NILC, \emph{Factsheet: Trump's Rescission of Protected Areas Policies...}, \href{https://www.nilc.org/resources/factsheet-trumps-rescission-of-protected-areas-policies-undermines-safety-for-all/\#:~:text=Factsheet\%3A\%20Trump\%27s\%20Rescission\%20of\%20Protected,churches\%2C\%20school\%2C\%20and\%20hospitals}{[Source Link]}. Epstein Becker & Green, \emph{The New Trump Administration's Immigration Enforcement Policy}, \href{https://www.ebglaw.com/insights/publications/the-new-trump-administrations-immigration-enforcement-policy-what-employers-must-know\#:~:text=The\%20New\%20Trump\%20Administration\%27s\%20Immigration,a\%20lobby\%20or\%20parking\%20lot}{[Source Link]}.}

\section*{Guantánamo Bay Plan and Broader Context}

\subsection*{Offshore Detention Plan}
In late January 2025, President Trump signed an executive order to potentially expand migrant detention capacity at \textbf{Guantánamo Bay (Gitmo), Cuba}, signaling plans for holding up to 30,000 individuals there.\footnote{CBS News, \emph{Trump administration sends first group...}, \href{https://www.cbsnews.com/news/trump-guantanamo-bay-migrant-detainees/\#:~:text=there\%20on\%20Tuesday}{[Source Link]} (Feb. 5, 2025); Center for Constitutional Rights (CCR), \emph{FOIA Request on Guantánamo EO}, \href{https://ccrjustice.org/sites/default/files/attach/2025/02/GTMO\%20EO\%20FOIA\%20FINAL\%202-5-25.pdf\#:~:text=the\%20Trump\%20Guant\%C3\%A1namo\%20EO\%20contemplates\%2C,threat\%20to\%20transfer\%20up\%20to}{[Source Link]} (Feb. 5, 2025) (mentioning 30,000 capacity); Reuters, \emph{Trump deporting people...}, \href{https://www.reuters.com/world/us/trump-set-broaden-arrests-deportation-routes-expand-immigration-crackdown-2025-02-21/\#:~:text=The\%20U,pushback\%20from\%20civil\%20liberties\%20groups}{[Source Link]} (Feb 21, 2025).} On February 5, 2025, the \textbf{first flight carrying migrant detainees apprehended on U.S. soil landed at Gitmo}.\footnote{CBS News, \emph{Trump administration sends first group...}, \href{https://www.cbsnews.com/news/trump-guantanamo-bay-migrant-detainees/\#:~:text=Homeland\%20Security\%2C\%20have\%20scrambled\%20to,implement\%20the\%20president\%27s\%20directive}{[Source Link]} (Feb. 5, 2025) (reporting flight with 10 detainees from Texas). DHS Secretary Kristi Noem later confirmed flights landed at Gitmo. CBS News, \emph{Court blocks Trump administration from deporting 3 Venezuelan immigrants...}, \href{https://www.cbsnews.com/news/3-venezuelan-immigrants-court-blocks-administration-deporting-guantanamo-bay/\#:~:text=The\%20Tren\%20de\%20Aragua\%20gang,S}{[Source Link]} (Mar. 21, 2025).} Detainees like Jhoan Bastidas reported being held in poor conditions in "holding tents" for extended periods (up to 16 days in some cases) with minimal provisions and contact.\footnote{AP, \emph{After a stint in Guantanamo Bay...}, \href{https://apnews.com/article/trump-guantanamo-migrants-venezuela-maduro-el-salvador-859984dd5231b5a7c22a7671ca646d48\#:~:text=MARACAIBO\%2C\%20Venezuela\%20,meals\%20that\%20left\%20him\%20hungry}{[Source Link]} (Mar. 25, 2025).} While initially framed for "high-threat" individuals,\footnote{CBS News, \emph{Trump administration sends first group...}, \href{https://www.cbsnews.com/news/trump-guantanamo-bay-migrant-detainees/\#:~:text=On\%20Tuesday\%20afternoon\%2C\%20a\%20plane,that\%20originated\%20in\%20Venezuela\%27s\%20prisons}{[Source Link]} (Feb. 5, 2025) (noting initial focus on alleged gang members).} reports and advocacy indicated non-violent or low-risk migrants were also transferred or targeted for transfer.\footnote{ACLU Press Release, \emph{Federal Appeals Court Keeps Block...}, \href{https://www.aclu.org/press-releases/federal-appeals-court-keeps-block-on-trump-use-of-alien-enemies-act-to-deport-immigrants\#:~:text=The\%20president\%E2\%80\%99s\%20use\%20of\%20this,judicial\%20review\%20or\%20due\%20process}{[Source Link]} (Mar. 26, 2025) (referencing plan to send individuals, including low-risk ones, to foreign prisons like Gitmo without due process).}

\subsection*{Legal Significance: Circumventing Habeas Corpus and Due Process}
Transferring detainees apprehended within the U.S. to Guantánamo is unprecedented\footnote{CCR, \emph{FOIA Request on Guantánamo EO}, \href{https://ccrjustice.org/sites/default/files/attach/2025/02/GTMO\%20EO\%20FOIA\%20FINAL\%202-5-25.pdf\#:~:text=Critically\%2C\%20no\%20one\%20has\%20ever,United\%20States\%E2\%80\%99\%20reputation\%20as\%20a}{[Source Link]} (Feb. 5, 2025) ("Critically, no one has ever been moved from U.S. soil to Guantánamo for detention...").} and widely viewed by legal experts as an attempt to place them outside the reach of U.S. constitutional protections, particularly the right to \textbf{habeas corpus} (to challenge detention in federal court, affirmed for Gitmo detainees in \textit{Boumediene v. Bush}) and Fifth Amendment due process.\footnote{Id.; ACLU Press Release, \emph{Federal Appeals Court Keeps Block...}, \href{https://www.aclu.org/press-releases/federal-appeals-court-keeps-block-on-trump-use-of-alien-enemies-act-to-deport-immigrants\#:~:text=\%E2\%80\%9CThe\%20decision\%20means\%20that\%20hundreds,Project\%2C\%20who\%20argued\%20the\%20case}{[Source Link]} (Mar. 26, 2025) (describing Gitmo as a "notorious black-hole prison... without any due process whatsoever"). A federal judge in New Mexico blocked the transfer of three Venezuelans to Gitmo specifically citing concerns about loss of access to counsel and courts. CBS News, \emph{Court blocks Trump administration...}, \href{https://www.cbsnews.com/news/3-venezuelan-immigrants-court-blocks-administration-deporting-guantanamo-bay/\#:~:text=It\%20asked\%20a\%20U,to\%20authorize\%20the\%20modest\%20injunction}{[Source Link]} (Mar. 21, 2025).} Such transfers, especially without clear statutory authorization under the INA, could be deemed \textit{ultra vires} and violate international prohibitions against arbitrary detention (e.g., ICCPR Article 9). This move, coupled with the attempted use of the Alien Enemies Act,\footnote{ACLU Press Release, \emph{Federal Appeals Court Keeps Block...}, \href{https://www.aclu.org/press-releases/federal-appeals-court-keeps-block-on-trump-use-of-alien-enemies-act-to-deport-immigrants\#:~:text=The\%20parties\%20were\%20in\%20court,and\%20protections\%20in\%20immigration\%20law}{[Source Link]} (Mar. 26, 2025).} signaled an effort to employ extraordinary measures for detention and removal, bypassing judicial review and established legal frameworks.\footnote{ACLU Press Release, \emph{Federal Appeals Court Keeps Block...}, \href{https://www.aclu.org/press-releases/federal-appeals-court-keeps-block-on-trump-use-of-alien-enemies-act-to-deport-immigrants\#:~:text=The\%20president\%E2\%80\%99s\%20use\%20of\%20this,judicial\%20review\%20or\%20due\%20process}{[Source Link]} (Mar. 26, 2025).}

\subsection*{Broader Policies}
These actions occurred alongside a nationwide expansion of \textbf{Expedited Removal} authority in January 2025, allowing ICE/CBP to deport undocumented individuals anywhere in the U.S. without an immigration judge hearing if they cannot prove two years' presence.\footnote{National Immigration Law Center (NILC), \emph{Know Your Rights: Expedited Removal Expansion}, \href{https://www.nilc.org/resources/know-your-rights-expedited-removal-expansion/}{[Source Link]} (Updated Jan. 2025). Note: While NILC source is undated in snippet, context places expansion around Jan 21, 2025.} This policy itself was flagged for likely due process violations due to the risk of erroneous removals without adequate review.\footnote{National Immigration Forum, \emph{Legislative Bulletin — Friday, January 31, 2025}, \href{https://immigrationforum.org/article/legislative-bulletin-friday-january-31-2025/\#:~:text=Experts\%20warn\%20that\%20such\%20quotas,individuals\%20are\%20detained\%20without}{[Source Link]} (Jan. 31, 2025) (reporting expert warnings).}

\section*{Legal and Civil Rights Response}

Civil liberties organizations (ACLU, CCR, NIJC, RFK Human Rights, etc.) responded rapidly with lawsuits, habeas petitions, motions to enforce prior settlements, and public condemnation.\footnote{See, e.g., ACLU Press Release, \emph{Court Rules Mahmoud Khalil’s Lawsuit...}, \href{https://www.aclu.org/press-releases/court-rules-mahmoud-khalils-lawsuit-challenging-his-unlawful-detention-by-ice-should-move-forward-in-new-jersey\#:~:text=In\%20the\%20early\%20morning\%20hours,Khalil\%E2\%80\%99s\%20attorneys\%20filed\%20a\%20petition}{[Source Link]} (Mar. 19, 2025) (Khalil lawsuit); ACLU of Illinois Press Release, \emph{22 People Arrested in ICE Raids...}, \href{https://www.aclu-il.org/en/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=The\%20motion\%20to\%20enforce\%20demands,forward\%20of\%20all\%20immigration\%20arrests}{[Source Link]} (Mar. 17, 2025) (Chicago raids motion); RFK Human Rights, \emph{Justice Roundup- March 2025}, \href{https://www.linkedin.com/pulse/justice-roundup-march-2025-rfkhumanrights-glnxe\#:~:text=The\%20Central\%20Louisiana\%20ICE\%20Processing,light\%20on\%20egregious\%20human\%20rights}{[Source Link]} (Mar. 2025) (Condemnation of detention center use).} Key legal arguments focus on violations of the First Amendment (retaliation against protected speech), Fourth Amendment (warrantless arrests and seizures), and Fifth Amendment (Due Process Clause violations – procedural and substantive – and potential Equal Protection issues related to targeting specific nationalities or viewpoints).\footnote{See, e.g., Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=\%E2\%80\%9CThe\%20government\%E2\%80\%99s\%20unlawful\%20policy\%20of,court\%20filing\%20on\%20March\%2013}{[Source Link]} (Mar. 27, 2025) (quoting Khalil's filing on unconstitutional targeting based on speech); ACLU of Illinois Press Release, \emph{22 People Arrested in ICE Raids...}, \href{https://www.aclu-il.org/en/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=,Mark\%20Fleming\%2C\%20associate\%20director\%20of}{[Source Link]} (Mar. 17, 2025) (arguing Fourth Amendment violations). Concerns also include violations of the INA, the Refugee Act, and international law principles like non-refoulement and the prohibition on arbitrary detention.} Courts have issued several TROs and injunctions, blocking specific deportations (Venezuelans via Alien Enemies Act, Suri),\footnote{Reuters, \emph{US appeals court upholds block...}, \href{https://www.reuters.com/world/us/us-homeland-secretary-visit-el-salvador-prison-holding-deported-venezuelans-2025-03-26/\#:~:text=Venezuelan\%20immigrants\%20under\%20a\%20little,18th\%20century\%20law}{[Source Link]} (Mar. 26, 2025); Al Jazeera, \emph{Who are the students...}, \href{https://www.aljazeera.com/news/2025/3/27/who-are-the-students-trump-wants-to-deport\#:~:text=However\%2C\%20the\%20Trump\%20administration\%20has,Semitism}{[Source Link]} (Mar. 27, 2025).} preventing detentions (Chung),\footnote{Reuters, \emph{Judge rules pro-Palestinian student...}, \href{https://www.reuters.com/legal/judge-rules-pro-palestinian-student-permanent-us-resident-cannot-be-detained-2025-03-25/\#:~:text=U,being\%20detained\%2C\%20court\%20records\%20showed}{[Source Link]} (Mar. 25, 2025).} blocking transfers to Guantánamo,\footnote{CBS News, \emph{Court blocks Trump administration...}, \href{https://www.cbsnews.com/news/3-venezuelan-immigrants-court-blocks-administration-deporting-guantanamo-bay/\#:~:text=It\%20asked\%20a\%20U,to\%20authorize\%20the\%20modest\%20injunction}{[Source Link]} (Mar. 21, 2025).} and returning venue (Khalil),\footnote{ACLU Press Release, \emph{Court Rules Mahmoud Khalil’s Lawsuit...}, \href{https://www.aclu.org/press-releases/court-rules-mahmoud-khalils-lawsuit-challenging-his-unlawful-detention-by-ice-should-move-forward-in-new-jersey\#:~:text=NEW\%20YORK\%20\%E2\%80\%93\%20The\%20Southern,absence\%20of\%20a\%20court\%20order}{[Source Link]} (Mar. 19, 2025).} suggesting judicial skepticism towards the administration's actions and justifications. Advocacy groups have characterized the overall pattern as authoritarian, McCarthyist, xenophobic, and having a dangerous chilling effect on dissent, noting the administration "blithely violate[s]... laws [and] ignore[s] court-approved [orders]" exhibiting "authoritarian and monarchical impulses."\footnote{See, e.g., The Guardian, \emph{‘A warning for students of color’...}, \href{https://www.theguardian.com/us-news/2025/mar/26/us-universities-students-israel-palestine-protests\#:~:text=Trump\%E2\%80\%99s\%20current\%20actions\%20against\%20pro,from\%20the\%20country\%2C\%E2\%80\%9D\%20she\%20said}{[Source Link]} (Mar. 26, 2025) (quoting experts comparing to past abuses and warning of chilling effect); NIJC, \emph{22 People Arrested...}, \href{https://immigrantjustice.org/press-releases/22-people-arrested-ice-raids-announce-federal-court-action-challenging-unlawful\#:~:text=Deportations}{[Source Link]} (Mar. 17, 2025) (quoting ACLU counsel).}

\section*{Conclusion}

In early 2025, U.S. immigration authorities under the Trump administration acted illegally and unconstitutionally. This includes the mass deportation of Venezuelan migrants without due process under a dubiously invoked, archaic wartime law (the Alien Enemies Act), carried out in blatant defiance of a federal court order attempting to halt it. A concurrent pattern involved targeting activists (particularly pro-Palestinian students, including LPRs) for detention and removal based on their protected speech, often using illegal tactics such as warrantless arrests, secretive transfers, and unsubstantiated national security claims, violating First, Fourth, and Fifth Amendment rights. Plans to utilize Guantánamo Bay for detaining migrants apprehended domestically, circumventing habeas corpus and judicial review, further underscore a trend towards lawlessness. These actions not only conflict with the U.S. Constitution and the Immigration and Nationality Act's procedural safeguards but also violate U.S. obligations under international refugee and human rights law. The deliberate circumvention of judicial oversight, use of emergency powers against specific groups, and targeting of dissent mirror patterns observed in authoritarian regimes.

\vspace{1em} % Add some vertical space before the source note


\end{document}